### 2024-04-23 parts/appendix.tex
move to appendix
move to appendix and clean up

@@

@@

>>>>

\AP
The following
\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
and is similar in shape to the \emph{finite counting automata} introduced by
\cite{SCHU62}.

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a function.
    The \intro{$\Rel$-polyregular function}
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is computed as follows:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is called \intro{$\Nat$-polyregular}.
    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
    When the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \intro{star-free $\Rel$-polyregular}
    (resp. \intro{star-free $\Nat$-polyregular}).
\end{definition}

\textbf{TODO: remove this example}
\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
    \proofref{size-of-word-nsf:ex}
\end{example}

\textbf{TODO: remove this lemma}
\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
    \label{transducer-nsf-npoly:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    Then, 
    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
        (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
        \kl{counter-free}).
\end{lemma}

\AP Let us briefly recall in the following lemmas alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide a nicer
syntax to construct examples. These characterizations were previously known,
and can be found for instance in \cite{DOUE23,CDTL23}. For every language $L
\subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^* \to \set{0,1}$ as the
indicator function of $L$. Given an $\MSO$ (resp. $\FO$) formula with
first-order free variables $\varphi(\vec{x})$, we define
$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
multiplication with elements $s \in \mathbb{S}$.

\begin{lemma}
    \label{polynomial-rational-polyreg:fact}
    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \NPoly$ (resp. $\ZPoly$);
        \item $f$ is a \kl{polyregular function} with output
            in $\set{1}^*$,
            post-composed with $\polysum$
            (resp. with output in $\set{-1,+1}^*$);
        \item $f$ has \kl{polynomial growth};
        \item \label{npoly-counting-mso:item} $f$ belongs to
            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
            with $\mathbb{S} = \Nat$
            (resp. $\mathbb{S} = \Rel$).
    \end{enumerate}
\end{lemma}

\begin{lemma}
    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \NSF$ (resp. $\ZSF$);
        \item $f$ is a \kl{star-free polyregular function} with output
            in $\set{1}^*$,
            post-composed with $\polysum$
            (resp. with output in $\set{-1,+1}^*$);
        \item $f$ belongs to
            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
            with $\mathbb{S} = \Nat$
            (resp. $\mathbb{S} = \Rel$).
    \end{enumerate}
\end{lemma}

%Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
%generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
%\begin{example}
%    \label{regular-language:ex}
%    Let $\Sigma$ be a finite alphabet, and
%    $L \subseteq \Sigma^*$. Then,
%    $L$ is a regular language if and only if
%    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
%    \kl{$\Nat$-polyregular function}.
%    Furthermore, $L$ is a star-free regular language
%    if and only if $\ind{L}$ is a
%    \kl{star-free $\Nat$-polyregular function}.
%\end{example}
%It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
%$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
%of first order free variables used to construct the function as in
%\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.

@@

\begin{proofof}[corrected-version:thm]
    The implications 
    \cref{corrected-4:item} $\implies$
    \cref{corrected-3:item} $\implies$
    \cref{corrected-2:item} are obvious.
    \cref{lem:correct-to-n-rat} proves
    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
    while \cref{n-rat-correct:lem}
    proves 
    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
    Note that the lemmas provide effective conversion procedures,
    and that \cref{lem:correct-to-n-rat} also provides a decision
    procedure.
\end{proofof}

@@

\begin{remark}
    \label{good-residual-ordering:fact}
    Let $k \in \Nat$, and let $f \colon \Sigma^* \to \Nat$. Then,
    $(\resleq{f}{k})$ is a quasi-ordering, satisfying the following
    extra properties:
    \begin{enumerate}
        \item For all $u,v,w \in \Sigma^*$, $u \resleq{f}{k} v$
            implies $uw \resleq{f}{k} vw$,
        \item If $u \resleq{f}{k} v$ and $v \resleq{f}{k} u$,
            then $\app{f}{u} = \app{f}{v}$.
    \end{enumerate}
\end{remark}

### 2024-04-23 parts/correction.tex
move to appendix and clean up

\AP When multiple indeterminates are involved, the notion of degree must be
adapted, and to ensure that the decision procedure is effective, one must
obtain explicit bounds on the \kl{translations} that are needed. To that end,
let us now introduce the ordering over which the induction of
\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
\kl{well-quasi-ordering}, and is in particular well-founded \cite[Hoare
quasi-ordering]{SCSC12}.

@@

\begin{theorem}
    \label{corrected-version:thm}
    Let $P \in \Rel[\vec{X}]$ be a polynomial.
    The following are equivalent:
    \begin{enumerate}
        \item \label{corrected-1:item} $P \in \CorrectPoly$,
        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
    \end{enumerate}
    Furthermore, the membership is decidable, and effective conversion
    procedures exist between all the representations.
\end{theorem}
\begin{proof}
    The implications 
    \cref{corrected-4:item} $\implies$
    \cref{corrected-3:item} $\implies$
    \cref{corrected-2:item} are obvious.
    \cref{lem:correct-to-n-rat} proves
    \cref{corrected-1:item} $\implies$ \cref{corrected-4:item},
    while \cref{n-rat-correct:lem}
    proves 
    \cref{corrected-2:item} $\implies$ \cref{corrected-1:item}.
    Note that the lemmas provide effective conversion procedures,
    and that \cref{lem:correct-to-n-rat} also provides a decision
    procedure.
\end{proof}

>>>>

\AP Let us now introduce the ordering over which the induction of
\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
\kl{well-quasi-ordering}, and is in particular well-founded \cite[Hoare
quasi-ordering]{SCSC12}.

@@

\begin{theorem}
    \label{corrected-version:thm}
    Let $P \in \Rel[\vec{X}]$ be a polynomial.
    The following are equivalent:
    \begin{enumerate}
        \item \label{corrected-1:item} $P \in \CorrectPoly$,
        \item \label{corrected-2:item} $P$ is \kl{represented} by a \kl{$\Nat$-rational series},
        \item \label{corrected-3:item} $P$ is \kl{represented} by a \kl{$\Nat$-polyregular function},
        \item \label{corrected-4:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
    \end{enumerate}
    Furthermore, the properties are decidable, and conversions effective.
\end{theorem}

### 2024-04-23 parts/prelim.tex
move to appendix and clean up

\begin{definition}
    Let $\mathcal{H}$ be a family of functions
    from $\Sigma^*$ to $\Nat$.
    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
    a tuple $(Q, q_0, \delta, \lambda, F)$ where
    \begin{itemize}
        \item $Q$ is a finite set of states,
        \item $q_0 \in Q$ is an initial state,
        \item $\delta \colon Q \times \Sigma \to Q$
            is a deterministic transition function,
        \item $\lambda \colon Q \times \Sigma \to \mathcal{H}$
            is a correction function,
        \item $F \colon Q \to \Nat$ is an output function.
    \end{itemize}
\end{definition}

\begin{definition}
    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
    The function
    $\aTransd \colon Q \times \Sigma^* \to \Nat$
    is defined inductively as follows:
        $\aTransd(q, \varepsilon) \defined F(q)$, and
        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
            + \lambda(q,a)(u)$.
    The function \intro{computed} by a transducer $\aTransd$
    is $w \mapsto \aTransd(q_0, w)$.
\end{definition}

\AP It is already known that this notion of transducers exactly captures $\NSF$
and $\NPoly$, and generalizes to outputs into other commutative semigroups than
$\Nat$. To make the above statement formal, let us recall that a
\intro{counter} in an automaton is a pair $u,w$ of words together with an
integer $n > 1$, such that $\delta(q_0, uw^n) = \delta(q_0, u)$ and
$\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$; where we used the
extension of $\delta \colon Q \times \Sigma \to Q$ to words by defining
$\delta(q, \varepsilon) = q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. An
automaton is \reintro{counter-free} if it contains no \reintro{counters}. This
notion can be lifted to \kl{$\mathcal{H}$-transducers} by considering counters
of their underlying automata.

\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
    \label{transducer-nsf-npoly:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    Then, 
    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
        (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
        \kl{counter-free}).
\end{lemma}

@@

\AP The \intro{growth rate} of a function $f \colon \Sigma^* \to \Rel$ is
defined as the minimal $k$ such that $\card{f(w)} =
\bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that the function
$f$ has \intro{polynomial growth}. We write $\reintro*\ZPoly[k]$ (resp.
$\reintro*\NPoly[k]$) for the functions in $\ZRat$ (resp. $\NRat$) that have
growth rate at most $k$.

It is clear that a function $f \in \ZPoly$ defined using a production function
$\pi \colon M^k \to \Rel$ has \kl{growth rate} at most $k-1$, and more
generally it is known that \kl{polyregular functions} have \kl{polynomial
growth} \cite{BOJA22}. Not all \kl{$\Nat$-rational series} have \kl{polynomial
growth}, a typical example being the function $a^n \mapsto 2^n$.

\begin{example}
    \label{size-of-word-nsf:ex}
    The map $f \colon w \mapsto \card{w} + 1$
    belongs to $\NSF$.
    \proofref{size-of-word-nsf:ex}
\end{example}

\AP
The following
\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
and is similar in shape to the \emph{finite counting automata} introduced by
\cite{SCHU62}.

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $M$ be a finite monoid, $\mu \colon \Sigma^* \to M$
    be a morphism, $k \in \Nat$, and 
    $\pi \colon M^k \to \Rel$ be a function.
    The \intro{$\Rel$-polyregular function}
    $\pi^\dagger \colon \Sigma^* \to \Rel$
    is computed as follows:
    \begin{equation*}
        \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_k} \pi(\mu(u_1), \dots, \mu(u_k))
        \quad .
    \end{equation*}
    When $\pi$ has co-domain $\Nat$, the function $\pi^\dagger$
    is called \intro{$\Nat$-polyregular}.
    We call $\pi$ the \emph{production function} of $\pi^\dagger$.
    When the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \intro{star-free $\Rel$-polyregular}
    (resp. \intro{star-free $\Nat$-polyregular}).
\end{definition}

\AP Let us briefly recall in the following lemmas alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide a nicer
syntax to construct examples. These characterizations were previously known,
and can be found for instance in \cite{DOUE23,CDTL23}. For every language $L
\subseteq \Sigma^*$ we define $\ind{L} \colon \Sigma^* \to \set{0,1}$ as the
indicator function of $L$. Given an $\MSO$ (resp. $\FO$) formula with
first-order free variables $\varphi(\vec{x})$, we define
$\vcount{\varphi(\vec{x})}$ as the function from $\Sigma^*$ to $\Nat$ that maps
a word $w$ to the number of assignments $\nu \colon \vec{x} \to w$, such that
$w, \nu \models \varphi(\vec{x})$. If $X$ is a set of functions from $\Sigma^*$
to $\Rel$, and a subset $\mathbb{S}$ of $\Rel$, we define
$\intro*\Span{\mathbb{S}}(X)$ as the smallest subset of functions from
$\Sigma^* \to \Rel$ containing $X$ and closed under (finite) sums and
multiplication with elements $s \in \mathbb{S}$.

\begin{lemma}
    \label{polynomial-rational-polyreg:fact}
    Let $f \in \NRat$ (resp. $f \in \ZRat$), then
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \NPoly$ (resp. $\ZPoly$);
        \item $f$ is a \kl{polyregular function} with output
            in $\set{1}^*$,
            post-composed with $\polysum$
            (resp. with output in $\set{-1,+1}^*$);
        \item $f$ has \kl{polynomial growth};
        \item \label{npoly-counting-mso:item} $f$ belongs to
            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \MSO})$
            with $\mathbb{S} = \Nat$
            (resp. $\mathbb{S} = \Rel$).
    \end{enumerate}
\end{lemma}

\begin{lemma}
    Let $f \in \NPoly$ (resp. $f \in \ZPoly$), then
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \NSF$ (resp. $\ZSF$);
        \item $f$ is a \kl{star-free polyregular function} with output
            in $\set{1}^*$,
            post-composed with $\polysum$
            (resp. with output in $\set{-1,+1}^*$);
        \item $f$ belongs to
            $\Span{\mathbb{S}}(\setof{\vcount{\varphi}}{\varphi(\vec{x}) \in \FO})$
            with $\mathbb{S} = \Nat$
            (resp. $\mathbb{S} = \Rel$).
    \end{enumerate}
\end{lemma}

%Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
%generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.
%\begin{example}
%    \label{regular-language:ex}
%    Let $\Sigma$ be a finite alphabet, and
%    $L \subseteq \Sigma^*$. Then,
%    $L$ is a regular language if and only if
%    $\ind{L} \colon \Sigma^* \to \set{0,1}$ is a
%    \kl{$\Nat$-polyregular function}.
%    Furthermore, $L$ is a star-free regular language
%    if and only if $\ind{L}$ is a
%    \kl{star-free $\Nat$-polyregular function}.
%\end{example}
%It has been proven in \cite{CDTL23,BOJA22} that the classes $\ZPoly[k]$ (resp.
%$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are also characterized by the maximal number
%of first order free variables used to construct the function as in
%\cref{npoly-counting-mso:item} of \cref{polynomial-rational-polyreg:fact}.

Let us briefly state that \kl{commutativity} is a decidable property of
\kl{$\Rel$-rational series}. The proof is straightforward in the case of
\kl{$\Rel$-polyregular functions}, and slightly more involved for general
\kl{$\Rel$-rational series}.

>>>>

\begin{definition}
    Let $\mathcal{H}$ be a family of functions
    from $\Sigma^*$ to $\Rel$.
    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
    a tuple $(Q, q_0, \delta, \lambda, F)$ where
         $Q$ is a finite set of states,
         $q_0 \in Q$ is an initial state,
         $\delta \colon Q \times \Sigma \to Q$
            is a deterministic transition function,
         $\lambda \colon Q \times \Sigma \to \mathcal{H}$
            is a correction function,
         and $F \colon Q \to \Rel$ is an output function.
\end{definition}

\begin{definition}
    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
    The function
    $\aTransd \colon Q \times \Sigma^* \to \Rel$
    is defined inductively as follows:
        $\aTransd(q, \varepsilon) \defined F(q)$, and
        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
            + \lambda(q,a)(u)$.
    The function \intro{computed} by a transducer $\aTransd$
    is $w \mapsto \aTransd(q_0, w)$.
\end{definition}

\AP Let us recall that a \intro{counter} in an automaton is a pair $u,w$ of
words together with an integer $n > 1$, such that $\delta(q_0, uw^n) =
\delta(q_0, u)$ and $\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$;
where we used the extension of $\delta \colon Q \times \Sigma \to Q$ to words
by defining $\delta(q, \varepsilon) = q$, and $\delta(q,au) =
\delta(\delta(q,a), u)$. An automaton is \reintro{counter-free} if it contains
no \reintro{counters}. This notion can be lifted to
\kl{$\mathcal{H}$-transducers} by considering counters of their underlying
automata.

\AP The notion of \kl{$\mathcal{H}$-transducer} can be used to inductively
define the classes $\intro*\NPoly[k]$, $\intro*\NSF[k]$, $\intro*\ZPoly[k]$ and
$\intro*\ZSF[k]$ as follows \cite[Theorems 5.15 and 7.10]{DOUE23}: in the base
case, $\reintro*\NPoly[-1] = \reintro*\NSF[-1]  = \reintro*\ZSF[-1] =
\reintro*\ZPoly[-1]$ is the singleton containing the constant function equal to
$0$; and for $k \geq 0$, $\reintro*\ZPoly[k]$ is the set of functions
\kl{computed} by a \kl{$\reintro*\ZPoly[k-1]$-transducer}, and
$\reintro*\NPoly[k]$ is the set of functions \kl{computed} by an
\kl{$\reintro*\NPoly[k-1]$-transducer} \emph{with an $\Nat$-valued output
function $F$}, and similarly for $\reintro*\NSF[k]$ and $\reintro*\ZSF[k]$,
where the transducer is furthermore required to be \kl{counter-free}. Then,
$\reintro*\NPoly = \bigcup_{k \in \Nat} \reintro*\NPoly[k]$, and classes
$\reintro*\NSF$, $\reintro*\ZPoly$ and $\reintro*\ZSF$ are defined similarly.
In the rest of the paper, we will always assume that
\kl{$\reintro*\NPoly[k]$-transducers} and \kl{$\reintro*\NSF[k]$-transducers}
have an $\Nat$-valued output functions, making a clean separation between
$\NPoly$ and $\ZPoly$.

@@

\AP One of the appeals of $\NPoly$ and $\ZPoly$ are the numerous
characterizations of these classes in terms of logic, weighted automata, and
the larger class of \kl{polyregular functions} \cite{CDTL23,DOUE23}. In this
paper, the main focus will be the connection to weighted automata, which is
based on the notion of \emph{growth rate}. The \intro{growth rate} of a
function $f \colon \Sigma^* \to \Rel$ is defined as the minimal $k$ such that
$\card{f(w)} = \bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that
the function $f$ has \intro{polynomial growth}. It turns out that for all $k
\in \Nat$, $\ZPoly[k]$ (resp. $\NPoly[k]$) are precisely functions in $\ZRat$
(resp. in $\NRat)$ that have growth rate at most $k$. Not all
\kl{$\Nat$-rational series} have \kl{polynomial growth}, a typical example
being the function $a^n \mapsto 2^n$.

Let us briefly state that \kl{commutativity} is a decidable property of
\kl{$\Rel$-rational series}, hence of \kl{$\Rel$-polyregular functions}. This
ensures that the present paper, focusing of \kl{commutative} functions,
considers functions in a decidable subclass. The proof is straightforward in
the case of \kl{$\Rel$-polyregular functions}, and slightly more involved for
general \kl{$\Rel$-rational series}.

### 2024-04-23 parts/starfree.tex
mise en forme
move to appendix and clean up

The goal of this section is to go beyond polynomials and consider the case of
\kl{commutative} \kl{$\Rel$-polyregular functions}. The hypothesis of
\kl{commutativity} reduces $\ZPoly$ to a combination of a finite number of
polynomials, arranged by counting modulo some numbers. This characterization
will be proven in \cref{decompose-polynomial:lem}, and will be the key
ingredient in the decision procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$
inside $\NPoly$. Note that \cref{decompose-polynomial:lem} is actually a
refinement of \cref{n-poly-combinatorics:lem} under the extra assumption of
\kl{commutativity}.

@@

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat$,
    and computable 
    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $w \in \Sigma^*$,
    \begin{equation*}
        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) 
        = P_{\moduloType[\omega](X_1, \dots, X_n)}
        \left(
            \floor{X_1/\omega}, \dots, \floor{X_n/\omega}
        \right)
        \quad .
    \end{equation*}
\end{lemma}

In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
of \emph{polynomial pumping arguments}
\cite{DOUE21,DOUE22,CDTL23}, we will provide a
semantic property in the form of \cref{k-combinatorial:def}, that is conjectured
to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.

@@

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and 
    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
    The following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
\end{theorem}

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, 
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{ultimately polynomial},
        \item $f \in \ZSF$,
        \item $f \in \NSF$.
    \end{enumerate}
    Furthermore, membership is decidable and conversions are effective.
\end{theorem}

>>>>

In this section, we leverage the decidability results of \cref{polynomials:sec}
to decide $\NPoly$ inside $\ZPoly$ and $\NSF$ inside $\NPoly$, both under the
extra assumption of \kl{commutativity}. The hypothesis of \kl{commutativity}
reduces $\ZPoly$ to a combination of a finite number of polynomials, arranged
by counting modulo some numbers. This characterization will be proven in
\cref{decompose-polynomial:lem}, and will be the key ingredient in the decision
procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
\cref{decompose-polynomial:lem} refines
\cref{n-poly-combinatorics:lem} for \kl{commutative} functions.

@@

\begin{lemma}
    \label{decompose-polynomial:lem}
    Let $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Rel$-polyregular function},
    where we fix the alphabet $\Sigma = \set{a_1, \dots, a_n}$.
    There exists a computable
    $\omega \in \Nat$,
    and computable 
    polynomials $\seqof{P_{t}}{t \in \ModuloTypes[\omega]^n}$
    such that for all $w \in \Sigma^*$,
    \begin{equation*}
        f\left(a_1^{X_1} \cdots a_n ^{X_n}\right) 
        = P_{\moduloType[\omega](X_1, \dots, X_n)}
        \left(
            \floor{X_1/\omega}, \dots, \floor{X_n/\omega}
        \right)
        \quad .
    \end{equation*}
    \proofref{decompose-polynomial:lem}
\end{lemma}

In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
of \emph{polynomial pumping arguments} \cite{DOUE21,DOUE22,CDTL23}, we will
provide a semantic property in the form of \cref{k-combinatorial:def}, that is
conjectured to characterize $\NPoly$ inside $\ZPoly$ in the general setting.

@@

\begin{theorem}
    \label{decidable-n-poly:thm}
    Let $k \in \Nat$, and 
    let $f \in \ZPoly[k]$ be a \kl{commutative} \kl{$\Rel$-polyregular function}.
    The following are equivalent:
    \begin{enumerate}
        \item \label{f-combinatorial:item} $f$ is \kl{$(k,\Nat)$-combinatorial},
        \item \label{f-npoly-combi:item} $f \in \NPoly[k]$,
    \end{enumerate}
    Furthermore, the properties are decidable,
    and conversions effective.
    \proofref{decidable-n-poly:thm}
\end{theorem}

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, 
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{ultimately polynomial},
        \item $f \in \ZSF$,
        \item $f \in \NSF$.
    \end{enumerate}
    Furthermore, membership is decidable and conversions are effective.
    \proofref{zsf-npoly-nsf:thm}
\end{theorem}
